%&latex
\documentclass{article}
\usepackage{csquotes}

\begin{document}

%+Title
\title{Social Finance}
\author{Jonathan Conning and Jonathan Morduch}\thanks{This is a very preliminary draft.  We would like to thank Leonardo Becchetti, Robert Lensink, and Ingmar Nyman for useful comments and discussions as well as to seminar participants at Tokyo University, George Washington University, Universit‡ di Roma Tor Vergata, and Ottawa University. We gratefully acknowledge research funding from the Ford Foundation in support of this project}
\date{\today}
\maketitle
%-Title

%+Abstract
\begin{abstract}
    We propose a framework for understanding how social investorsí who seek to maximize a combination of private and social returns from investments in a portfolio of new and established microfinance institutions.  The model takes into account the endogeneity of loan contract terms as well as the capital structure of the financial institutions that may emerge to serve target groups of borrowers differentiated primarily by their levels of initial average net worth, and how social investments might transform those patterns. We build upon one of the workhorse models of modern corporate finance (Tirole, 2007) which features limited liability, multiple layers of moral hazard and costly monitoring to explain patterns of financial intermediation, and in our framework, the role and modes of social investment.  We pinpoint the role of the subsidies and guarantees implicit in social investorsí equity and quasi-equity investments and the role they play in attracting private capital investors and sustaining productivity-enhancing financial intermediation that might otherwise not have taken place.  
\end{abstract}
%-Abstract


\section{Introduction}
The working modes of philanthropy and business are blurring in two ways. First, philanthropists are increasingly using methods borrowed from for-profit investors. The strategies of venture capitalists are being adapted by ìventure philanthropistsî seeking to fund socially-minded institutions with promising ideas at early stages of development. They hope that the ideas will form the basis of sustainable institutions capable of serving populations at wide scale. As a recent survey of philanthropy put it:  

\begin{displayquote}
The modern philanthropist dislikes the term charity, preferring to speak about his social investments. His language is entrepreneurial, sprinkled with references to metrics, scalability, leverage, and venture philanthropy (Weinsberg, 2006).  
\end{displayquote}

The changes are expanding to foundations with the increasing favor of program-related investments over grants (PRIs include loans, equity shares, recoverable grants, linked deposits, and guarantees); parallel changes are afoot within government agencies (Marble, 1989). 

Oikocredit, for example, a Dutch social investment organization, has since 1978 been making long-term loans to ìenterprises and institutions for the empowerment of poor people in developing countries .î Their first ìsocial investmentî in microfinance, for example, was a \$150,000 loan for 5 years at 9\% interest to Alternativa, a non-governmental organization providing microcredit to street vendors and small-scale entrepreneurs in Lima, Peru. Between 1994 and the end of 2004, Oikocredit made a cumulative total of 285 loans or investment agreements to 161 institutions in 41 countries, totaling \$88 million. Their net profit, returned to shareholders as a dividend, was 2\%. The dividend allowed investors a modest return on capital, but they earned substantially less than alternative investments. Their relative loss is their implicit charitable contribution (Heinen, 2005). 

The second turn is that many socially-minded institutions receiving funding also increasingly look like businesses. Museums have raised entry prices and run ambitious shops with mail-order catalogues. Non-profit schools are hiking tuitions.  Environmental groups buy and sell land. Community development programs work with banks to facilitate mortgages in low-income neighborhoods. Indeed, some of these institutions are forming commercially viable entities and competing in the marketplace on the strength of their balance sheets alone. Opportunity International, a US-based network of microfinance institutions, owns a for-profit insurance company. BRAC, the largest NGO in Bangladesh, owns a for-profit bank. Chicagoís Shorebank, the largest community development bank in the United States, is a holding company composed of non-profit and for-profit entities. BASIX has adapted the Shorebank model to South India. 
?

The microfinance sector is home to some of the best-known examples of the new hybrids, and is in many ways the leading edge. Muhammad Yunus, the 2006 Nobel Peace Prize winner and founder of the Grameen Bank, a financial institution for the poor of Bangladesh, created a ìsocial businessî funded by social investors. 

We tackle the logic and tensions inherent in social finance -- óthe support, with philanthropic objectives, of nonprofits, social businesses like the Grameen Bank, and profit-maximizing businesses serving the poor. The principles behind the new world of philanthropy and social action have not been well-explored by economists. We develop a theory of ìsocial financeî to parallel the modern theory of corporate finance. 

Most generally, we assess the justification for the subsidy inherent in social investment. We go further to describe a framework that describes how subsidy can be deployed for optimal impact. A focus is on using funds to build institutions that can achieve scale and attract additional revenue sources and investments. Leverage is critical, and we describe two important kinds of leverage. One, having to do with mobilizing locally-held capital, is often overlooked and is highlighted below. 

Our consideration of effective social investment begins with two ideas. The first is that effective giving can lead to multiplier effects. It can ìcrowd inî other resources by catalyzing the investments of others, increasing the recipientís ability to ìleverageî its resources. Leverage (i.e., to be able to obtain funds to use in addition to an institutionís own equity) is critical as it determines an institutionís ability to reach wide scale. Effective social investment can be used to expand leverage and hence market coverage, holding constant other social concerns. 

The second consideration is that philanthropic giving can change the incentives for recipients to work hard (as can commercial lending). 

In this framework, profitability is neither necessary nor sufficient for attracting commercial capital. Contrary to some popular claims, earning profits ñ even high profits ñwill frequently not expand the leverage of microfinance to take those institutions to wide scale. The presence of a social investor can however tip the scales for private resources to pour in. Total output may be enhanced by supporting institutions that would not exist if profitability was the sole concern.  Of course it is no great surprise that a subsidized investment can expand an activity or a sector, the claim however is that social investors and philanthropic capital can expand output and welfare by repairing and expanding markets. 

The framework has a specific implication for the most effective use of social investment. We describe arguments that push social investment either ìhighî or ìlowî when directed to poor communities. Specifically, we show that smart subsidy involves allocations that either go to institutions serving relatively well-off (but still poor) customers, where the potential for leverage is maximized, or to institutions serving the poorest customers. There is less ìbang for the buckî in supporting a middle range of institutions that are self-sufficient with their own resources but for which leverage (and thus potential scale) is limited. 


\section{Two layers of moral hazard}
The analysis focuses on social investment in microfinance institutions. The particular gap filled by social investors is created by the unwillingness of most commercial banks to serve the poor. One well-known difficulty for banks stems from moral hazard on the part of borrowers. In practice, though, moral hazard is mitigated by microfinance institutions through the development of technologies, contracts, and management practices that limit risk. All of this takes effort, but the high repayment rates achieved by leading practitioners are testimonies to impressive successes. The essential insight in mitigating moral hazard is that incentives are improved by exposing people to the consequences of their actions. In the microfinance context, the denial of new loans in the case of default (via ìprogressive lendingî) and the peer pressure associated with group lending can reduce moral hazard by adding serious repercussions for bad outcomes (Armendariz de Aghion and Morduch, 2005). 

Given the success in controlling moral hazard, investors ought to be interested in funding microfinance institutions and expanding their scale. Investments should be pouring in, and they are starting, but most investments are made by social investors with explicit philanthropic objectives. 
While social investors are willing to forego the kind of financial returns that alternative investments could bring, they nonetheless work hard to preserve their capital. The result is that most money channeled through microfinance investment funds and via International Financial Institutions supports ìtop-tier,î mature, financially self-sufficient institutions (Abrams and von Stauffenberg, 2007).  The growing, not-yet-profitable institutions that are needed to expand the sector are targeted by less than one fifth of the funds in the MicroBanking Bulletinís survey of funds (Barres, 2005:53). 
One of the chief difficulties for investorsówhether driven by commercial or social endsñis a second layer of moral hazard. The basic moral hazard problem described above involves the microfinance institution and the customer. This second problem involves the investor and the microfinance institution. Once it is the investorís resources that are ultimately at risk (rather than just the microfinance institutionís resources or the customersí), the incentives are reduced for the institution to be diligent in managing finances, monitoring loans, and innovating. The incentives to properly screen potential customers weaken, as do the incentives to chase down delinquent borrowers.  
The risks are not trivial. Oikocredit, for example, as of late 2004 faced a portfolio at risk (90-day) of 7 percent on its \$67 million social investment portfolio (Heinen 2005: 33).  In a review of 104 loan contract disbursed before the end of 2002, the average repayment rate was about 91 percent. Seventy four percent of contracts were found to be repaid ìalmost always on time,î but ì15 percent require continuous monitoring to follow overdue payments closely, and another 11 percent can be classified as bad-debt cases that are persistent in their default. The latter cases come with high monitoring and legal costs and are likely to lead to (partial) loan loss in the end.î The challenges among the African institutions are greater: ìOnly 57 percent of MFIs in the African sample manage to pay ìon time,î while more than one out of four are structurally delayed in payment of invoiced amounts (Heinen 2005: 34-35).î  The worst performers tended to newly-started, medium-sized urban banks or NGOs in Africa or Asia (Heinen 2005: 39). 
Without collateral fully backing many of the institutionís financial portfolios, this kind of risk becomes a major concern for prudent investors. Recognizing moral hazard matters since, under moral hazard, potential asset-poor customers typically face credit rationingóeven though they may have worthwhile investment projects. The projects are worthwhile in the sense that they would generate enough income to allow borrowers to in principle pay back loans with interest, and an entrepreneur would profitably undertake the project with their own money if they had the necessary fundsThe potential trouble is that the borrowers cannot find funding for their projects. The same holds for worthy institutions that are rationed out of the capital market due to information problems.  Much of the modern theory of financial intermediation has been built upon this particular insight (Diamond 1984; Tirole, 2006) but the application of this theory to understanding the role of public policy and social investors in shaping the emergent financial intermediation landscape remains considerably unexplored. 
We argue that commercial investors play important roles, but by virtue of their philanthropic objectives, social investors can help increase leverage, expand scale, and increase total output in a sector in ways that commercial investors by themselves often cannot. 
Specifically, we demonstrate cases where philanthropic-giving can in principle raise output, increasing an economyís overall productivityóand lead to GDP increases in addition to poverty reduction. The possibility arises because social investors, by definition, are willing to absorb costs in order to bring gains. This is, of course, their reason for existence. They can do more than shift resources: they can also help increase the total sum of resources available. The question then is: How can they most powerfully enact such ìproductivity-enhancingî investments.  
Our consideration of effective social investment begins with two ideas. The first consideration is that philanthropic giving can expand the contract space by being willing to subsidize the ëlimited liability rentsí that often stand in the way of more trades in the commercial loan market.  Limited liability rents refer to the returns that must be left with borrowers, or the loan officers that monitor them so as to provide incentives to effort when limited liability constraints restrict the incentives that can be provided by threatening punishments for example by using legal mechanisms to seize collateral. When such rents plus the cost of loan monitoring become too large relative to expected project gains, lenders may walk away from making loans to individuals without collateral wealth, leaving socially valuable projects unfunded.  A small subsidy in such contexts can a times be enough to turn a small private loans that might otherwise have been unprofitable now profitable. If enough of those small loans become profitable, a new microfinance institution might emerge where otherwise it might not have. 
The second consideration is that effective giving may lead to multiplier effects.  It can ìcrowd inî other resources by catalyzing the investments of others, increasing the recipientís ability to ìleverageî its resources. Leverage (i.e. to be able to obtain funds to use in addition to an institutionís own equity) is critical as it determines an institutionís ability to reach wide scale, and wider scale can lower average fixed costs.  Effective social investment aims to expand leverage, holding constant other social concerns. 
In this framework, profitability is neither necessary nor sufficient for attracting commercial capital. Merely earning profit will not generate the leverage to take institutions to wide scale. Moreover, productivity can be enhanced by supporting institutions that would not exist if profitability was the sole concern. 
The framework has a specific implication for the most effective use of social investment. We describe arguments that push social investment either ìhighî or ìlowî when directed to poor communities. Specifically, we show that smart subsidy involves allocations that either go to institutions serving relatively well-off (but still poor) customers, where the potential for leverage is maximized, or to institutions serving the poorest customers. There is less ìbang for the buckî in supporting a middle range of institutions that are self-sufficient with their own resources but for which leverage (and thus potential scale) is limited. 


\subsection{Customizing Class Options}
Select `Insert', `Document Properties ...', the `Generic' tab
and then modify desired class options in appeared dialog.
Changes will be applied after pressing the 'OK' button.

%+Bibliography
\begin{thebibliography}{99}
\bibitem{Label1} ...
\bibitem{Label2} ...
\end{thebibliography}
%-Bibliography

\end{document}


